---
title: "Sales Trend Visualization"
output:
  word_document: default
  html_document: default
---

# Sales Trend Visualization Project

## 1. Install & Load Libraries
```{r}
#install.packages(c("dplyr", "ggplot2"))
library(dplyr)
library(ggplot2)
```

---

## 2. Load Dataset
```{r}
data <- read.csv("train.csv")

cat("DATASET OVERVIEW\n")
cat("Rows:", nrow(data), "\n")
cat("Columns:", ncol(data), "\n\n")

str(data)
head(data)
```

---

## 3. Missing Values Before Cleaning
```{r}
cat("MISSING VALUES BEFORE CLEANING\n")
missing_before <- colSums(is.na(data))
print(missing_before[missing_before > 0])
```

**Insight:** Dataset contains missing values that need cleaning.

---

## 4. Data Cleaning
```{r}
#  Convert Date properly
data$Order.Date <- as.Date(data$Order.Date, format="%d-%m-%Y")

#  Clean Sales column
data$Sales <- gsub(",", "", data$Sales)   # remove commas
data$Sales <- gsub("\\$", "", data$Sales) # remove $
data$Sales <- as.numeric(data$Sales)      # convert to numeric

#  Create Month + Year properly (IMPORTANT FIX)
data$Year <- format(data$Order.Date, "%Y")
data$Month <- format(data$Order.Date, "%b")
data$Month_Num <- as.numeric(format(data$Order.Date, "%m"))  # ⚠️ numeric banana zaruri

#  Check NA values
cat("NA in Sales:", sum(is.na(data$Sales)), "\n")
cat("NA in Date:", sum(is.na(data$Order.Date)), "\n")

#  Remove NA rows (clean dataset)
data <- data[!is.na(data$Sales) & !is.na(data$Order.Date), ]

cat("Final Rows after cleaning:", nrow(data), "\n")
```

---

## 5. Feature Engineering
```{r}
data$Year <- format(data$Order.Date, "%Y")
data$Month <- format(data$Order.Date, "%b")
data$Month_Num <- as.numeric(format(data$Order.Date, "%m"))  
```

---

## 6. Q: What is the total sales?
```{r}
total_sales <- sum(data$Sales)
cat(" Total Sales:", total_sales)
```

---

## 7. Q: How does sales change month-by-month?
```{r}
monthly <- data %>%
  group_by(Month, Month_Num) %>%
  summarise(Sales = sum(Sales, na.rm = TRUE), .groups="drop") %>%
  arrange(Month_Num)

ggplot(monthly, aes(x = factor(Month, levels = Month), y = Sales, group=1)) +
  geom_line(color="blue", size=1.2) +
  geom_point(color="red", size=3) +
  labs(title="Monthly Sales Trend",
       x="Month", y="Sales") +
  theme_minimal()
peak_month <- monthly$Month[which.max(monthly$Sales)]
cat("\n Highest Sales Month:", peak_month)

```
Highest Sales Month: Feb

---

## 8. Q: Which months contribute the most?
```{r}
ggplot(monthly, aes(x = reorder(Month, Sales), y = Sales)) +
  geom_bar(stat="identity", fill="skyblue") +
  coord_flip() +
  labs(title="Monthly Sales Contribution",
       x="Month", y="Sales") +
  theme_minimal()

top2 <- head(monthly[order(-monthly$Sales), ], 2)
cat("\n Top Months:", paste(top2$Month, collapse=", "))
```

**Insight:** Top Months: Feb, Aug

---


## 9. Q: How does sales change year-wise?
```{r}
yearly <- data %>%
  group_by(Year) %>%
  summarise(Sales = sum(Sales, na.rm = TRUE), .groups="drop")

ggplot(yearly, aes(x=Year, y=Sales, group=1)) +
  geom_line(color="green", size=1.3) +
  geom_point(size=3) +
  labs(title="Yearly Sales Trend",
       x="Year", y="Sales") +
  theme_minimal()

best_year <- yearly$Year[which.max(yearly$Sales)]
cat("\n👉 Best Performing Year:", best_year)
```

**Insight:** Identifies the year with highest business performance.
             Best Performing Year: 2018

---

## 10. Q: Which region generates the highest sales?
```{r}
region_sales <- data %>%
  group_by(Region) %>%
  summarise(Total_Sales = sum(Sales), .groups = "drop")

print(region_sales)

top_region <- region_sales[which.max(region_sales$Total_Sales), ]


ggplot(region_sales, aes(x = Region, y = Total_Sales)) +
  geom_bar(stat = "identity" , fill="#008080") +
  ggtitle("Region-wise Sales") +
  xlab("Region") +
  ylab("Sales")

cat("Top region:", top_region$Region)

```

**Insight:** Shows which geographical region generates highest revenue.
             Top region: West

---

## 11. Q: Which category has the highest sales?
```{r}
category <- data %>%
  group_by(Category) %>%
  summarise(Sales = sum(Sales, na.rm = TRUE), .groups="drop") %>%
  arrange(desc(Sales))

ggplot(category, aes(x=reorder(Category, Sales), y=Sales)) +
  geom_bar(stat="identity", fill="orange") +
  coord_flip() +
  labs(title="Sales by Category",
       x="Category", y="Sales") +
  theme_minimal()

cat("\n Top Category:", category$Category[1])
```

**Insight:** Identifies most profitable product category. 
             Top Category: Technology


---

## 12. Q: What are the top 10 sub-categories?
```{r}
subcat <- data %>%
  group_by(Sub.Category) %>%
  summarise(Sales = sum(Sales, na.rm = TRUE), .groups="drop") %>%
  arrange(desc(Sales)) %>%
  head(10)

ggplot(subcat, aes(x=reorder(Sub.Category, Sales), y=Sales)) +
  geom_bar(stat="identity", fill="pink") +
  coord_flip() +
  labs(title="Top 10 Sub-Categories",
       x="Sub Category", y="Sales") +
  theme_minimal()

cat("\n Top Sub-Category:", subcat$Sub.Category[1])
```
**Insight:** Top Sub-Category: Phones


---



## 13. Final Conclusion
```{r results='asis'}

# Best Month
best_month <- monthly$Month[which.max(monthly$Sales)]

# Best Category
best_category <- category$Category[1]

# Best Sub-category
best_sub <- subcat$Sub.Category[1]

# Best Region (if exists)
if(exists("region")) {
  best_region <- region$Region[1]
} else {
  best_region <- "Not Available"
}

cat(paste0(
"📌 Final Insights:

- Highest sales month: ", best_month, "
- Sales show an overall increasing trend over time
- Top category: ", best_category, "
- Top sub-category: ", best_sub, "
- Top region: ", best_region
))
```