---
title: "Sales Trend Analysis"
output:
  word_document: default
  html_document:
    df_print: paged
---

# Sales Trend Visualization Project

## 1. Install & Load Libraries
```{r}
#install.packages(c("dplyr", "ggplot2"))
library(dplyr)
library(ggplot2)
```

---

## 2. Load Dataset
```{r}
data <- read.csv("train.csv")

cat("DATASET OVERVIEW\n")
cat("Rows:", nrow(data), "\n")
cat("Columns:", ncol(data), "\n\n")

str(data)
head(data)
```

---

## 3. Missing Values Before Cleaning
```{r}
cat("MISSING VALUES BEFORE CLEANING\n")
missing_before <- colSums(is.na(data))
print(missing_before[missing_before > 0])
```

**Insight:** Dataset contains missing values that need cleaning.

---

## 4. Data Cleaning
```{r}
data$Order.Date <- as.Date(data$Order.Date, format="%m/%d/%Y")
data <- na.omit(data)
```

---

## 5. Feature Engineering
```{r}
data$Year <- format(data$Order.Date, "%Y")
data$Month <- format(data$Order.Date, "%m")
```

---

## 6. Missing Values After Cleaning
```{r}
cat("MISSING VALUES AFTER CLEANING\n")
print(colSums(is.na(data)))
```

**Insight:** Dataset is now clean and ready for analysis.

---

## 7. Monthly Sales Analysis
```{r}
monthly_sales <- data %>%
  group_by(Year, Month) %>%
  summarise(Total_Sales = sum(Sales), .groups = "drop")

print(monthly_sales)
```

**Insight:** Sales fluctuate across months indicating seasonal trends.

---

## 8. Monthly Sales Trend Graph
```{r}
monthly_sales$Date <- as.Date(paste(monthly_sales$Year, monthly_sales$Month, "01", sep="-"))

ggplot(monthly_sales, aes(x = Date, y = Total_Sales)) +
  geom_line() +
  ggtitle("Monthly Sales Trend") +
  xlab("Time") +
  ylab("Sales")
```

**Insight:** Shows overall sales growth and fluctuations over time.

---

## 9. Year-wise Sales
```{r}
yearly_sales <- data %>%
  group_by(Year) %>%
  summarise(Total_Sales = sum(Sales), .groups = "drop")

print(yearly_sales)

best_year <- yearly_sales[which.max(yearly_sales$Total_Sales), ]

cat("Best performing year:", best_year$Year)
```

```{r}
ggplot(yearly_sales, aes(x = Year, y = Total_Sales)) +
  geom_bar(stat = "identity", fill="lightgreen") +
  ggtitle("Year-wise Sales") +
  xlab("Year") +
  ylab("Sales")
```

**Insight:** Identifies the year with highest business performance.

---

## 10. Region-wise Sales
```{r}
region_sales <- data %>%
  group_by(Region) %>%
  summarise(Total_Sales = sum(Sales), .groups = "drop")

print(region_sales)

top_region <- region_sales[which.max(region_sales$Total_Sales), ]

cat("Top region:", top_region$Region)
```

```{r}
ggplot(region_sales, aes(x = Region, y = Total_Sales)) +
  geom_bar(stat = "identity", fill="skyblue") +
  ggtitle("Region-wise Sales") +
  xlab("Region") +
  ylab("Sales")
```

**Insight:** Shows which geographical region generates highest revenue.

---

## 11. Category-wise Sales
```{r}
category_sales <- data %>%
  group_by(Category) %>%
  summarise(Total_Sales = sum(Sales), .groups = "drop")

print(category_sales)

top_category <- category_sales[which.max(category_sales$Total_Sales), ]

cat("Top category:", top_category$Category)
```

```{r}
ggplot(category_sales, aes(x = Category, y = Total_Sales)) +
  geom_bar(stat = "identity", fill="pink") +
  ggtitle("Category-wise Sales") +
  xlab("Category") +
  ylab("Sales")
```

**Insight:** Identifies most profitable product category.

---

## 12. Final Conclusion
```{r}
cat("FINAL INSIGHTS SUMMARY\n")
cat("1. Sales show seasonal fluctuations.\n")
cat("2. Business performance varies year-wise.\n")
cat("3. Certain regions generate higher revenue.\n")
cat("4. Some categories dominate sales.\n")
```